---
title: "An example R Markdown notebook"
author: "Zakhodiakin G.V., postlogist@gmail.com"
date: "06.09.2019"
output: 
  html_document: 
    toc: true
    toc_float: true
---

# Introduction

In contrast to a script, an R Markdown Document may contain both code and text, graphics, mathematical formulae, and the code's output. Such documents are stored as a plain text files with .Rmd extension.
When loaded in RStudio, R Markdown Documents are displayed as interactive notebooks consisting of text and code chunks, for example:

```{r Loading Packages, message=FALSE, warning=FALSE}

library(tidyverse)

```


You can execute the code chunks step-by-step (`Ctrl-Shift-Enter`), or run them all at once (see the `Run` button's menu in RStudio).


```{r Loading and viewing a built-in dataset}

data(mpg)
head(mpg)

```

The results are displayed just after the code chunk creating them - very cool!

Similarly, text ouptut and plots are displayed just after the code chunk.

```{r Visualization of the relationship and calculation of Persons r}

# Data Visualization
qplot(x = hwy, y = cty, data = mpg)

# Computing the corellation coefficient
cor.test(~ cty + hwy, data = mpg) 

```


```{r Building and evaluating a regression model}

# Model building
m_cty <- lm(cty ~ hwy, data = mpg)

# Viewing model summary
summary(m_cty) 

```

The calclulated results can be embedded in a text: 

The model's $R^2$ is `r round(summary(m_cty)$r.squared, 3)`.

Let's visualize the model we've built.

```{r Visualizing the model}

# Getting the model coefficients
coef_m_cty <- coef(m_cty)

# Plotting data and the trend line
qplot(x = hwy, y = cty, data = mpg) +
  geom_abline(intercept = coef_m_cty[1], 
              slope = coef_m_cty[2],
              colour = "red") +
  labs(title = "City and highway mileage",
       y = "City mileage, miles per gallon",
       x = "Highway mileage, miles per gallon")
```

In contrast to scipts, the working directory for a markdown document is always the directory containing this R Markdown document. Thus we don't care about setting the working directory.
As a best practice, always store all analysis-related files in the same folder, as the R markdown document that uses them, or in some sub-folders.

```{r Saving a plot to file}

ggsave("example_plot_2.png")

```

In R Studio we can see the output in the same window, as the R Markdown Document. But despite this, the output is not stored in the Rmd file. This significantly simplifies working with Rmd files via source control systems, like Git/GitHub. The changes in files are displayed in a human readable form.


When you need to send you analysis results to another person, you can convert, or 'Knit' the R Markdown document into a HTML web-page or a Word document, that will contain all the code, text and output. The resulting file is self-contained and can be displayed even when your recipient doesn't have R on his or her computer. Or even phone.

To compile the report, press the `Knit` button on the RStudio toolbar.




# An example data analysis using data from the Internet

R can load data from local files, as well as from remote sources, such as web servers. In this example, we'll download a time series from the [Sophist database](http://sophist.hse.ru) maintainted by HSE.

To run this example, we'll need `forecast` and `sophisthse` packages, that are included in the R Portable distribution prepared for this class. If you did install R using standard installers, you'll probably need to install these packages. In this case, uncomment and run the following chunk once:

```{r Installing Packages}
#install.packages("forecast")
#install.packages("sophisthse")
```

Don't forget to comment or remove the lines above, when the packages are installed. This needs to be done just once.


Loading the packages (usually this chunk will be at the top of the document)

```{r Loading the packages}
library(forecast)
library(sophisthse)
```


Downloading the time series data (Retail Turnover in Russia)

```{r Loading the RTRD_M series}
retail <- sophisthse('RTRD_M')
head(retail)
sophisthse_metadata(retail)
```

Let's keep only the data for the retail turnover in monetary units (RTRD_M)` and see, how the time series looks like.

```{r Plotting the time series}

rtrd_m <- retail[, 'RTRD_M']
autoplot(rtrd_m)

```


The historical period is very deep. Let's keep just the data since 2010.


```{r Selecting data}

rtrd_subset <- window(rtrd_m, start = c(2010, 1)) #Beginning Year and Month 

```

Let's examine the seasonal structure for time series.

```{r Time series decomposition}

autoplot(decompose(rtrd_subset, type = 'multiplicative'))

```

One can see, that after the 2014's crisis the growth rate has declined.

Now we'll calculate the forecast.
As a first step, a model should be fit using historical data. By default, it will be selected automatically from the exponential smoothing family of models. 

```{r Building and viewing the time series model}

m_rtrd_subset <- ets(rtrd_subset)
summary(m_rtrd_subset)

```

Now we have a model and can use it for forecasting. Let's use a 24 month forecasting horizon.

```{r Forecasting}

f_rtrd_subset <- forecast(m_rtrd_subset, h = 24)
autoplot(f_rtrd_subset) +
  labs(title = "Predicted retail turnover in Russia, Bln. Rub",
       x = NULL,
       y = NULL,
       fill = "95% Prediction\ninterval")

```


