---
title: "Exploratory data analysis of alcohol consumption by secondary school students"
author: <put your name here>
output: 
  html_document: 
    toc: true
    toc_float: true
---

# Introduction

In this assignment you'll practice your data visualization skills by performing an exploratory analysis. Your objective is to find factors correlated with the alcohol consumption by secondary school students. 

The dataset was collected by Paul Cortez and Alice Silva by administering a questionnaire to students of two portugese schools. Basic social and demographic data as well as performance of each student were collected. Later the dataset was submitted to UCI machine learning repository by Fabio Pagnotta and Hossain Mohammad Amran. Information about the dataset is available [here](https://archive.ics.uci.edu/ml/datasets/STUDENT+ALCOHOL+CONSUMPTION). You are advised to take a quick look on the paper linked on that page to get familiar with the data collection methodology and the interpretation by original authors.

For use in this assignment, a subset of the original data was selected. It is available in the [student-mat.xlsx](student-mat.xlsx) file. The file contains data on a mathematics class students. Codes for categorical variables were replaced by text labels to simplify interpretation.

## Attribute information:

- `school` - student's school
- `sex` - student's sex (`F` - female or `M` - male) 
- `age` - student's age (numeric: from `15` to `22`) 
- `address` - student's home address type (`urban` or `rural`) 
- `famsize` - family size (`3 or less`, `4 or more`) 
- `Pstatus` - parent's cohabitation status (`living together` or `living apart`) 
- `Medu` - mother's education (ordered: from `0 - none` to `4 - higher education`)
- `Fedu` - father's education (same as `Medu`)
- `Mjob` - mother's job (nominal: `teacher`, `health` care related, civil `services` (e.g. administrative or police), `at_home` or `other`) 
-  `Fjob` - father's job (same as `Mjob`) 
- `reason` - reason to choose this school (nominal: close to `home`, school `reputation`, `course` preference or `other`) 
- `guardian` - student's guardian (nominal: `mother`, `father` or `other`) 
- `traveltime` - home to school travel time (ordered: `1 - <15 min.`, `2 - 15 to 30 min.`, `3 - 30 min. to 1 hour`, or `4 - >1 hour`) 
- `studytime` - weekly study time (ordered: `1 - <2 hours`, `2 - 2 to 5 hours`, `3 - 5 to 10 hours`, or `4 - >10 hours`) 
- `failures` - number of past class failures (numeric: n if 1<=n<3, else `4`) 
- `schoolsup` - extra educational support (yes or no) 
- `famsup` - family educational support (yes or no) 
- `paid` - extra paid classes within the course subject (Math) (yes or no) 
- `activities` - extra-curricular activities (yes or no) 
- `nursery` - attended nursery school (yes or no) 
- `higher` - wants to take higher education (yes or no) 
- `internet` - Internet access at home (yes or no) 
- `romantic` - with a romantic relationship (yes or no) 
- `famrel` - quality of family relationships (ordered: from `1 - very bad` to `5 - excellent`) 
- `freetime` - free time after school (ordered: from `1 - very low` to `5 - very high`) 
- `goout` - going out with friends (same as `freetime`) 
- `Dalc` - workday alcohol consumption (same as `freetime`) 
- `Walc` - weekend alcohol consumption (same as `freetime`) 
- `health` - current health status (ordered: from `1 - very bad` to `5 - very good`) 
- `absences` - number of school absences (numeric: from `0` to `93`) 
- `G1` - first period grade (numeric: from `0` to `20`) 
- `G2` - second period grade (same as `G1`) 
- `G3` - final grade (same as `G1`) 


## Assignment

Build data visualizations to answer study questions listed below using ggplot2. Interpret your findings regarding each question and write them down after each plot. All plots should have a descriptive title and clear labeling for axes and legends. 

**Tips:**

- You may find helpful the chapter on [graphics for communication](http://r4ds.had.co.nz/graphics-for-communication.html) in R for Data Science book. 

- You may also want to take a look on the [Exploratory data analysis](http://r4ds.had.co.nz/exploratory-data-analysis.html) chapter in that book.

- Please note, that most variables in the dataset are categorical. Some of them, like `sex` or `Medu` are nominal (unordered). Some have a particular order of levels (`goout`, `health`). You are advised to read this [blog post](https://www.r-bloggers.com/choosing-colour-palettes-part-ii-educated-choices/) on selecting a color palette for a particular nature of data. An interactive guide to selecting a suitable palette is available on the [colorbrewer](http://colorbrewer2.org/) website. Read about each palette's purpose there by clicking the `i` icons on that page. You can adjust the color scale with `scale_color_brewer()` and `scale_fill_brewer()` functions.

- For information on text formatting using Markdown, see this [reference](http://rmarkdown.rstudio.com/authoring_basics.html) and [this cheat sheet](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf).

# Instructions on how to submit your solution

1. Download this Rmarkdown file to your computer. Note, that it requires a data file - `students-mat.xlsx` to reside in the same folder as this .Rmd file.

2. Put your plotting code in a code chunk below each question. Write a short interpretation of your findings in text just after the output of the code chunk.

3. When finished all questions, remove all sections of the document not relevant to your analysis (like 'Task', 'Instructions on how to submit...')

4. Create a HTML report by pressking `Knit` button in RStudio. If your document has errors, this process will fail. You can review the errors using the `R Markdown` tab in the lower left part of RStudio window. If you can't see it, try to expand the section of RStudio labelled `Console` (see the lower left part).

5. Send me the resulting html document.

Remember to put your name(s) in the `author:` field on top of this document.


# Analysis

Put your code to load libraries and the data in the following code chunk. You'll need the `tidyverse` and the `scales` packages for this assignment. I've added the readxl package which is required to read in the Excel file containing the dataset. Use `read_excel()` to read-in the data file. 

```{r Load libraries}
# Your code here
library(readxl)

```


## Question 1: Can we analyze the students of both schools together?

The dataset contains survey data from two schools in Portugal. To check, if the results can be analyzed as a combined dataset, compare the social and demographical characteristics of students in the two schools. 

### How many observations do we have from each school?

```{r Count observations by school}
# Your code here

```

**Conclusion:** ...

### Are the age distributions similar across the schools?

```{r Compare the age distributions}
# Your code here

```
 
**Conclusion:** ...


### Are the proportions of girls and boys similar across the schools?

```{r Gender proportions}
# Your code here

```

**Conclusion:** ...

### Are the proportions of students from urban and rural locations similar across the schools?

```{r Home address types}
# Your code here

```

**Conclusion:** ...

### Are the students' family sizes similar across the schools?

```{r Family sizes across the schools}
# Your code here

```

**Conclusion:** ...

### Are the parents' education levels similar across the schools?

```{r Parents education levels}
# Your code here

```

**Conclusion:** ...

### Are the parents' occupation similar across the schools?

```{r Parents occupation across the schools}
# Your code here

```

**Conclusion:** ...

### Are the reasons to choose the school similar across the schools?

```{r Reasons to choose the school}
# Your code here

```

**Conclusion:** ...

### Are the alcohol usage levels similar across the schools?

Compare weekend and workday alcohol consumption across the schools.

```{r Alcohol consumption across the schools}
# Your code here

```

**Conclusion:** ...

**Note:** you can observe that workday alcohol consumption is much less prevalent for both schools. Please, use the weekend alcohol consumption (`Walc`) for further analysis to ensure reliability of conclusions due to larger sample size.

### Is the relationship between alcohol consumption and the student's performance similar across the schools?

Compare the distributions and the average final grades (`G3`) for different levels of alcohol use (`Walc`) across schools.

```{r Alcohol consumption and final grades}
# Your code here

```

**Conclusion:** ...

You can observe differences in both characteristics of students and their performance across the schools. Both schools show quite unexpected patterns for relationship between alcohol consumption and performance. Looks like the [Ballmer's peak](http://observer.com/2012/04/bottoms-up-the-ballmer-peak-is-real-study-says/) is real for mathematics as well.

For consistency, take data from school with the largest number of observations - `Gabriel Pereira` for further analysis. User `filter()` function from the `dplyr` package to filter the data.

**Note:** the following code assumes you have loaded data into the `alco` variable. Adjust variable name according to what you've used.

```{r Subset data}
alco_gp <- filter(alco, school == 'Gabriel Pereira')
```


## Question 2: Is alcohol use correlated with student's performance?

You've already seen the relationship between the alcohol consumption and the grades. Use appropriate plots to explore relationships between alcohol use and other performance variables: the number of previous class failures and absenteeism.

```{r Alcohol use and performance}
# Your code here


```

**Conclusion:** ...

## Question 3: Is alcohol use correlated with student's health?

Visualize the reltaionship between student's alcohol use and health.

```{r Alcohol use and health}
# Your code here

```

**Conclusion:** ...

## Question 4: What social factors are correlated with alcohol use?

Use appropriate plots to study the correlation between social factors and alcohol use.

### Gender and age

```{r Gender, age and alcohol use}
# Your code here


```


**Conclusion:** ...

### Place of living (address type)

```{r Place of living}
# Your code here

```

**Conclusion:** ...

### Family size and parents' cohabitation status

```{r Family size and parents cohabitation}
# Your code here



```

**Conclusion:** ...


### Parents' occupation

```{r Parents occupation}
# Your code here


```

**Conclusion:** ...


### Social relationships

What kind of social relationships is correlated with alcohol use the most: family relationships (`famrel`), romantic relationship (`romantic`) or friendship (`goout`)?


```{r Social relationships and alcohol use}
# Your code here

```

**Conclusion:** ...

## Question 5: What corrective actions are the most efficient?

Compare the distribution of final grades (`G3`) for each corrective action: extra educational support (`schoolsup`), family educational support (`famsup`), extra paid Math classes (`paid`), extra-curricular activities (`activities`).

```{r Corrective actions and grades}
# Your code here

```

**Conclusion:** ...
