---
title: "regression"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=F, warning=F}
library(tidyverse)
library(GGally) # for scatterplot matrices
library(modelr) # for convenient models manipulation 
```


```{r}
data(mpg)
head(mpg)
```


Exploring correlations using the scatterplot matrix (the quick easy way):

```{r, message=F, warning=F}
mpg %>%
  select(displ, cty, cyl, hwy) %>%
  ggpairs()
```

Scatterplot matrices are too cluttered for datasets with lots of columns.


The tidyverse way for plotting is to "melt" all the predictor columns into one and to use ggplot with facets:

```{r}
mpg %>%
  select(displ, cty, cyl, year, hwy) %>% 
  gather(key = 'parameter', value = 'value', displ:cyl) %>% 
  ggplot(mapping = aes(x = value, y = hwy)) + 
    geom_point(aes(color = factor(year))) +
    geom_smooth(method = 'lm', se = FALSE) +
    facet_wrap(~ parameter)
  
```

If you want numerical values of Pearson's correlation coefficient - use built-in `cor()` function.

It works for any two vectors (columns):

```{r}
cor(mpg$cty, mpg$hwy)
```

```{r}
with(mpg, cor(cty, hwy))
```

Or for the entrire table. Note: you should remove all non-numerical and non-relevant columns before using `cor()`.

```{r}
mpg %>%
  select(displ, cty, cyl, year, hwy) %>%
  cor()
```


# The First Model

Use `lm()` to estimate a linear model. The formula notation is used to specify the model's formula: `y ~ X`

```{r}
#classical way
m1 <- lm(hwy ~ cty, data = mpg)

#using pipe
m1b <- 
  mpg %>%
  lm(hwy ~ cty, data = .)

```

Now the model is stored now in the `m1` variable. Try to devise a clear naming system for your models. It will simplify things a lot when you'll build many models.


```{r}
summary(m1)
```

Compare the model's residual standard error to the standard deviation of `hwy`:

```{r}
sd(mpg$hwy)
```

We see improvement against the case when no information on the predictor `cty` is used.

How to extract coefficients:

```{r}
coef(m1)['cty']
```

You can do so also in text:

The slope coefficient is `r coef(m1)['cty']`

# Forecasting

The built-in `predict()` function returns a vector of point forecasts. The `newdata` argument is used to specify the data to apply the model to. This is a regular table, which must contain all the variables used for the model. Just `cty` in our case.

Applying the model to the same data, to get the fitted values:

```{r}
predict(m1, newdata = mpg) %>% head(10)
```

Applying the model to new data:

```{r}

newd <- tibble(cty = c(10, 20, 30))
newd

predict(m1, newdata = newd)
```


You can add predictions to the table using `modelr::add_predictions()` function:

```{r}
newd %>%
  add_predictions(m1)
```

Or, in a classical way, just `cbind()` the vector of predicted values and the original data:

```{r}
newd %>%
  cbind(predict(m1, newdata = newd))
```



This is how you can add the interval forecast - the confidence and the prediction intervals:


```{r}
m1 %>%
  predict(newdata = mpg,
          interval = 'prediction',
          level = .8) %>%
  head()
```

This is not a table, but a matrix. To use tidyverse's functions, convert to a table.


```{r}

confint_m1 <- predict(m1, newdata = mpg, 
                      interval = 'confidence',
                      level = .8)   %>%
  as_tibble() %>%
  rename(lci = 'lwr', uci = 'upr')


predint_m1 <- predict(m1, newdata = mpg, 
                      interval = 'prediction',
                      level = .8) %>%
  as_tibble() %>%
  rename(lpi = 'lwr', upi = 'upr') %>%
  select(-fit)

forecast_mpg <- 
  cbind(mpg, confint_m1, predint_m1)

```

And, finally, plotting the forecast:

```{r}
forecast_mpg %>%
  ggplot(aes(x = cty)) +
  geom_ribbon(aes(ymin = lpi, ymax = upi),
              fill = 'lightskyblue') +
  geom_ribbon(aes(ymin = lci, ymax = uci),
              fill = 'lightgray') +
  geom_line(aes(y = fit), color = 'red') +
  geom_point(aes(y = hwy)) +
  ylab('hwy')
```

